---
title: "Exploratory analysis of compositional data (part II)"
format: 
  html:
    code-fold: false
    code-tools: true
bibliography: references.bib
editor: visual
---

## Exploratory analysis of a microbial data set

Here we use the "88 soils" data set [@lauber2009pyrosequencing] containing bacterial communities in 88 soils from across North and South America.

### Load packages and data data

```{r}
#| label: load_data
#| warning: false
#| message: false
#| echo: true

library(phyloseq)
library(dplyr)
library(ggplot2)

soil <- readRDS("data/soil_processed/soil.RDS")
```

Take a look at the data

```{r}
#| label: look_at_data
#| warning: false
#| echo: true

soil

# Read count table
otu_table(soil)[1:10, 1:6]

# Taxonomy table
taxtab <- tax_table(soil)
head(taxtab)

# Sample data
sampdata <- sample_data(soil)
head(sampdata)
```

### Microbial composition

We first have a look at the microbial compositions on phylum level.

```{r}
soil_phyla <- tax_glom(soil, taxrank = "Phylum")
```


Stacked bar plot of the observed "absolute" abundances:

```{r}
plot_bar(soil_phyla, fill = "Phylum")
```

Stacked bar plot of the relative abundances:

```{r}
# Compute relative abundances
otutab <- otu_table(soil_phyla)
otutab <- sweep(otutab, 1, FUN = `/`, rowSums(otutab))*100
soil_phyla@otu_table@.Data <- otutab
```

```{r}
plot_bar(soil_phyla, fill = "Phylum")
```

### Distance calculation and visualization

Distance functions provided by the phyloseq package:

```{r}
dist_methods <- unlist(distanceMethodList)
dist_methods
```

#### MDS plot with Bray Curtis distance

```{r}
soil_mds <- ordinate(soil, method = "MDS", distance = "bray")
```

First a plot without coloring metadata.

```{r}
mds_soil <- plot_ordination(soil, soil_mds, title = "MDS of 88 soil data")

mds_soil + theme_bw() + theme(text = element_text(size = 14)) + geom_point(size = 3)
```

One could identify two main clusters, separated by the first axis. 

We now build two pH categories (acidic and basic) and color the samples in the MDS plot accordingly.

```{r}
# Add categorical variable to the data frame 
sampdata$ph_type <- as.factor(ifelse(sampdata$ph < 6.7, "acid", "basic")) 

head(sampdata)

table(sampdata$ph_type)
```
Add the new data frame to the phyloseq object.

```{r}
sample_data(soil) <- sampdata
```


```{r}
mds_soil <- plot_ordination(soil, soil_mds, color = "ph_type",
                            title = "MDS of 88 soil data colored by pH type")

mds_soil + theme_bw() + theme(text = element_text(size = 14)) + geom_point(size = 3)
```

Indeed, pH type separates samples along Axis.1. 

#### MDS plots for several distance measures

Now, we apply all available distance functions to the data, compute the corresponding MDS embeddings, and plot them.

First, we have to remove the unwanted distances.

```{r}
filtered_methods <- dist_methods[!dist_methods %in% c("unifrac", "wunifrac",
                                                      "dpcoa", "ANY", "morisita")]
```

Now we can apply these distance metrics to our data. 

```{r}
gather_ordination_data <- function(d_method, phylo_obj){
  # Calculate the MDS matrix using the distance
  ordinate_obj <- ordinate(phylo_obj, method = "MDS", distance = d_method)
  # Get the data for plotting
  plot_data <- plot_ordination(phylo_obj, ordinate_obj)$data
  # Add a column for distance
  plot_data$distance <- d_method
  return(plot_data)
}

mds_plot_data <- bind_rows(lapply(filtered_methods, gather_ordination_data,
                                  phylo_obj = soil))

head(mds_plot_data)
```

We color again according to pH type.

```{r}
#| fig-width: 14
#| fig-height: 9

mds_soil <- ggplot(mds_plot_data, aes(x=Axis.1, y=Axis.2, color=ph_type)) +
  geom_point() +
  facet_wrap(~ distance, scales = "free") +
  labs(x="Axis.1",
       y="Axis.2",
       title="Separation of 88 soil samples according to different distances")

mds_soil + theme_bw() + theme(text = element_text(size = 12))
```

