---
title: "Practical applications of learned concepts in R"
format: 
  html:
    code-fold: false
    code-tools: true
bibliography: references.bib
editor: visual
---

## Network learning and analysis

Here, we learn how a microbial association network is constructed and analyzed using the R package NetCoMi.

::: {.callout-tip collapse="false"}
### The R package NetCoMi

The [NetCoMi](https://github.com/stefpeschel/NetCoMi) [@peschel2021netcomi] package is specifically designed to **construct, analyze, and compare networks** for microbiome data. It implements the complete workflow of learning and analyzing microbial networks explained in lecture 8. Instead of using several functions for each of the steps, `NetCoMi` provides a single function for network construction (`netConstruct()`), so the package streamlines the workflow considerably. The user can choose from a variety of methods for **data preprocessing, association estimation, sparsification, and transformation**. The returned `microNet` object can then be passed to `netAnalyze()` (the network analysis function) so that all necessary information is available for the network analysis workflow.
:::

### Install and load packages

We start by installing NetCoMi from GitHub. Two of NetCoMi's dependencies are also only available on GitHub. We will install them manually beforehand to avoid problems.

```{r}
#| label: install_netcomi
#| eval: false

devtools::install_github("zdk123/SpiecEasi")

devtools::install_github("GraceYoon/SPRING")

devtools::install_github("stefpeschel/NetCoMi", force = TRUE, ref = "TSE",
                         dependencies = c("Depends", "Imports", "LinkingTo"),
                         repos = c("https://cloud.r-project.org/",
                                   BiocManager::repositories()))
```

Load packages needed in this exercise.

```{r}
#| label: load_packages
library(NetCoMi)
library(phyloseq)
```

### Load the data

The 88 soil dataset [@lauber2009pyrosequencing] is again used in this example.

```{r}
#| label: load_data
#| warning: false
#| echo: true

soil <- readRDS("data/soil_processed/soil.RDS")
```

Take a look at the data

```{r}
#| label: look_at_data
#| warning: false
#| echo: true

soil

# Read count table
otu_table(soil)[1:10, 1:6]

# Taxonomy table
taxtab <- tax_table(soil)
head(taxtab)

# Sample data
sampdata <- sample_data(soil)
head(sampdata)
```

### Rename genera

If we want to generate a network on genus level, we run into two problems:

-   There are unidentified genera.
-   Not all genera are unique.

```{r}
#| label: taxtab
table(taxtab[, "Genus"])
```

Solutions:

-   Unknown genera are named by higher taxonomic ranks.
-   Duplicates are numbered to make them unique.

NetCoMi provides a function `renameTaxa()`, which does this job for us. It can also be used to rename the taxa according to a desired pattern.

```{r}
#| label: rename_genera
# Rename taxonomic table and make Rank6 (genus) unique
soil_renamed <- renameTaxa(soil, 
                           pat = "<name>", 
                           substPat = "<name>_<subst_name>(<subst_R>)",
                           numDupli = "Genus")

taxtab_renamed <- tax_table(soil_renamed)

head(cbind(taxtab[, "Genus"], taxtab_renamed[, "Genus"]), 15)
```
### Network construction

Now, we can pass the phyloseq object with renamed genera to NetCoMi's `netConstruct()` function.

SpiecEasi is used in the "MB" mode (Meinshausen & Bühlman neighborhood selection) to measure associations between the genera.

A few comments on the arguments:

-   The network is created on genus level.
-   The data are filtered as follows:
    -   Only samples with a total of at least 100 reads are included (argument `filtSamp`).
    -   Only taxa occuring in at least 5 samples are included (argument `filtTax`).
-   `measure` defines the association or dissimilarity measure, in our case SpiecEasi. Additional arguments are passed to `spiec.easi()` via `measurePar`. We reduce the number of StARS repetitions to 10 to reduce runtime.
-   Normalization and zero handling is done internally by `spiec.easi()`, so we set `normMethod` and `zeroMethod` to "none".
-   `sparsMethod` is also set to "none" (SpiecEasi includes StARS model selection to obtain a sparse network).
-   The "signed" method is used to transform associations into dissimilarities (argument `dissFunc`). Thus, strongly negatively associated taxa have a high dissimilarity and conversely a low similarity, corresponding to low edge weights on the network plot.

```{r}
#| label: netconstruct_single
#| eval: false
net <- netConstruct(soil_renamed,
                    taxRank = "Genus",
                    filtSamp = "totalReads",
                    filtSampPar = list(totalReads = 100),
                    filtTax = "numbSamp",
                    filtTaxPar = list(numbSamp = 5),
                    measure = "spieceasi",
                    measurePar = list(method = "mb",
                                      pulsar.params = list(rep.num = 10)),
                    zeroMethod = "none",
                    normMethod = "none",
                    sparsMethod = "none", 
                    dissFunc = "signed",
                    verbose = 3,
                    seed = 12345)
```

```{r}
#| label: save_single_net
#| echo: false
#| eval: false
saveRDS(net, file = "data/networks/spieceasi_net.RDS")
```

```{r}
#| label: read_single_net
#| echo: false
net <- readRDS("data/networks/spieceasi_net.RDS")
```

`netConstruct()` returns an object of the class `microNet`, which contains all matrices generated during network construction.

The object also contains an edge list, giving each edge's estimated association, dissimilarity, and adjacency. Let's take a quick look at the edges with the highest and lowest edge weights:

```{r edge_list}
#| label: edge_list
edgelist <- net$edgelist1[order(net$edgelist1$adja, decreasing = TRUE), ]
head(edgelist)
tail(edgelist)
```

### Network analysis

The `microNet` object is now passed to `netAnalyze` to perform network analysis.

A few comments on the arguments:

* `centrLCC = TRUE`: Centralities are calculated only for nodes in the largest connected component (LCC). 
* Clusters: Identified using greedy modularity optimization (by `cluster_fast_greedy` from [`igraph`](https://igraph.org/r/) package).
* Hubs: Nodes with an eigenvector centrality value above the empirical 95% quantile of all eigenvector centralities in the network.
* `weightDeg = FALSE` and `normDeg = FALSE`: The degree of a node is simply the number of its adjacent nodes.
* `gcmHeat = FALSE`: By default, a heatmap of the Graphlet Correlation Matrix (GCM) is plotted, which we disable here (`gcmHeat = FALSE`).

```{r}
#| label: netanalyze_single
netprops <- netAnalyze(net, 
                       centrLCC = TRUE,
                       clustMethod = "cluster_fast_greedy",
                       hubPar = "eigenvector",
                       hubQuant = 0.95,
                       weightDeg = FALSE,
                       normDeg = FALSE,
                       gcmHeat = FALSE)
```

```{r}
#| label: summary_single
summary(netprops)
```

### Highlight node properties

In the first plot, **node colors** represent the detected **clusters** and **node sizes** are scaled by **eigenvector centrality**. **Hub nodes** are highlighted by default. Singletons are not included in the plot. 

Note that nodes are sometimes placed too close together so that the labels overlap. You may need to play around with the `repulsion` argument until you find a value where the labels are legible, but also the clusters are still well recognizable.

```{r, fig.width=13, fig.height=10}
#| label: netplot_soil_props
#| fig-width: 13
#| fig-height: 10
plot(netprops,
     repulsion = 0.95,
     rmSingles = TRUE,
     shortenLabels = "none",
     labelScale = TRUE,
     nodeSize = "eigenvector",
     nodeSizeSpread = 3,
     nodeColor = "cluster", 
     hubBorderCol = "gray40",
     cexNodes = 1.8,
     cexLabels = 3,
     edgeTranspHigh = 20,
     title1 = "Network properties highlighted", 
     showTitle = TRUE,
     cexTitle = 2.3,
     mar = c(1, 3, 4, 8))

legend(0.7, 1.1, cex = 1.7, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)
```

### Highlight data features

We now **color** nodes according to their **phylum**. The **node sizes** are proportional to a taxon's sum of **clr-transformed abundances**. 

```{r, fig.width=15, fig.height=10}
#| label: netplot_soil_phyla
#| fig-width: 13
#| fig-height: 10
#| 
# Generate vector with phylum names for node coloring
phyla <- as.factor(taxtab_renamed[, "Phylum"])
names(phyla) <- taxtab_renamed[, "Genus"]

# Reorder levels
phyla <- reorder(phyla, phyla, FUN = length, decreasing = TRUE)

# Create color vector
colvec <- rainbow(length(levels(phyla)))

plot(netprops,
     repulsion = 0.95,
     rmSingles = TRUE,
     shortenLabels = "none",
     labelScale = TRUE,
     nodeSize = "clr",
     nodeColor = "feature", 
     featVecCol = phyla, 
     colorVec =  colvec,
     highlightHubs = FALSE,
     cexNodes = 1.2,
     cexLabels = 3,
     edgeTranspHigh = 20,
     title1 = "Data features highlighted", 
     showTitle = TRUE,
     cexTitle = 2.3,
     mar = c(1, 10, 4, 6))

# Add legends
legend(0.7, 1.1, cex = 1.7, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)

# Colors used in the legend should be equally transparent as in the plot
col_transp <- colToTransp(colvec, 60)

legend(-1.8, 1.1, cex = 1.7, pt.cex = 2.5, title = "Phylum:", 
       legend=levels(phyla), col = col_transp, bty = "n", pch = 16) 
```

### Select Proteobacteria

If we are interested in a sub-network (here only genera belonging to proteobacteria), we can also filter the nodes in the network plot.

```{r, fig.width=15, fig.height=10}
#| label: netplot_soil_proteo
#| fig-width: 15
#| fig-height: 10

sel <- names(phyla[phyla == "Proteobacteria"])
sel <- sel[sel %in% colnames(net$adjaMat1)]

plot(netprops,
          nodeFilter = "names",
     nodeFilterPar = sel,
     repulsion = 0.95,
     rmSingles = TRUE,
     shortenLabels = "none",
     labelScale = TRUE,
     nodeSize = "eigenvector",
     nodeSizeSpread = 3,
     nodeColor = "cluster", 
     hubBorderCol = "gray40",
     cexNodes = 1.8,
     cexLabels = 3,
     edgeTranspHigh = 20,
     title1 = "Sub-network for proteobacteria", 
     showTitle = TRUE,
     cexTitle = 2.3,
     mar = c(1, 3, 4, 8))

legend(0.7, 1.1, cex = 1.7, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)
```

## Network comparison (acidic vs. basic)

### Network construction

Now, we might be interested in **how the association networks differ** between samples with a high and a low pH, or in other words, **between basic and acidic samples**. 

```{r}
#| label: split_phylo
# Split the phyloseq object into two groups
soil_acid <- phyloseq::subset_samples(soil_renamed, ph < 6.7)
soil_basic <- phyloseq::subset_samples(soil_renamed, ph > 6.8)

soil_acid
soil_basic
```

```{r}
#| label: nsamp
n_basic <- phyloseq::nsamples(soil_basic)
```

We filter the `r n_basic` samples (sample size of the smaller group) with highest frequency to make the sample sizes equal and thus ensure comparability.

```{r}
#| label: netconstruct_groups
#| eval: false
net_groups <- netConstruct(data = soil_acid,
                           data2 = soil_basic,
                           taxRank = "Genus",
                           filtSamp = "highestFreq",
                           filtSampPar = list(highestFreq = n_basic),
                           filtTax = "totalReads",
                           filtTaxPar = list(totalReads = 1),
                           measure = "spieceasi",
                           measurePar = list(method = "mb",
                                             pulsar.params = list(rep.num = 10)),
                           zeroMethod = "none",
                           normMethod = "none",
                           sparsMethod = "none", 
                           dissFunc = "signed",
                           verbose = 3,
                           seed = 12345)
```

```{r}
#| label: saverds_groups
#| echo: false
#| eval: false
saveRDS(net_groups, file = "data/networks/spieceasi_netcomp.RDS")
```

```{r}
#| label: readrds_groups
#| echo: false
net_groups <- readRDS("data/networks/spieceasi_netcomp.RDS")
```

### Network analysis

All network measures are now computed for both networks. 

```{r}
#| label: netanalyze_groups
netprops_groups <- netAnalyze(net_groups, 
                              centrLCC = TRUE,
                              clustMethod = "cluster_fast_greedy",
                              hubPar = "eigenvector",
                              hubQuant = 0.95,
                              weightDeg = FALSE,
                              normDeg = FALSE,
                              gcmHeat = FALSE)
```

### Visual comparison

We start with a visual comparison of the two networks using NetCoMi's plot function. 

```{r, fig.width=17, fig.height=8}
#| label: netplot_acid_basic_difflay
#| fig-width: 17
#| fig-height: 8

plot(netprops_groups,
     repulsion = 0.96,
     rmSingles = TRUE,
     labelScale = FALSE,
     nodeSize = "eigenvector",
     nodeSizeSpread = 2,
     nodeColor = "cluster", 
     sameColThresh = 2,
     hubBorderCol = "darkgray",
     cexNodes = 2,
     edgeTranspHigh = 20,
     title1 = "Acidic", 
     title2 = "Basic",
     showTitle = TRUE,
     cexTitle = 2,
     cexLabels = 0.8,
     mar = c(1, 4, 4, 4))

# Overlay a transparent plot on which the legend is plotted
par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), mar=c(0, 0, 0, 0), new=TRUE)
plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')

legend(-0.2, -0.9, cex = 1.5, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)
```

Layout computed separately for each network -> difficult to visually compare certain associations

We now use the same layout in both groups.

```{r, fig.width=17, fig.height=8}
#| label: netplot_acid_basic_samelay
#| fig-width: 17
#| fig-height: 8

plot(netprops_groups,
     sameLayout = TRUE,
     layoutGroup = "union",
     repulsion = 0.96,
     rmSingles = "inboth",
     labelScale = FALSE,
     nodeSize = "eigenvector",
     nodeSizeSpread = 2,
     nodeColor = "cluster", 
     sameColThresh = 2,
     hubBorderCol = "darkgray",
     cexNodes = 2,
     edgeTranspHigh = 20,
     title1 = "Acidic", 
     title2 = "Basic",
     showTitle = TRUE,
     cexTitle = 2,
     cexLabels = 0.8,
     mar = c(1, 4, 4, 4))

# Add legend
par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), mar=c(0, 0, 0, 0), new=TRUE)
plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
legend(-0.2, -0.8, cex = 1.7, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"), 
       bty = "n", horiz = TRUE)
```


### Quantitative network comparison

```{r}
#| label: netcomp
netcomp <- netCompare(netprops_groups, 
                      permTest = FALSE,
                      nPerm = 1000,
                      cores = 6,
                      seed = 12345,
                      storeAssoPerm = TRUE,
                      fileStoreAssoPerm = "general/network_data/spring_assoPerm",
                      verbose = TRUE)
```

```{r}
#| label: summary_netcomp
summary(netcomp, 
        groupNames = c("Acidic", "Basic"),
        numbNodes = 5)
```



