---
title: "Practical applications of learned concepts in R"
format: 
  html:
    code-fold: false
    code-tools: true
editor: visual
---

## Example

TBD

Please run the code provided to replicate some of the analyses. Make sure you can explain what all the analysis steps do and that you understand all the results.

In addition, there are some extra tasks (`Task 1`), where no R code is provided. Please do these tasks when you have time available at the end of the lab.

### Load the data

Read the data, and convert to matrix format.

```{r}
#| label: load-data
#| warning: false
#| echo: true

mrna <- read.table("data/data_example.txt", header=T, sep="\t", dec=".")

# Convert to matrix format

mrna <- as.matrix(mrna)
```

Print the data

```{r}
#| label: look-at-data
#| warning: false
#| echo: true

mrna[1:4, 1:4]
```

Visualize the overall distribution of expression levels by histogram

```{r}
hist(mrna, nclass=40, xlim=c(-5,5), col="lightblue")
```

::: callout-note
## Task 1

*This is a callout-note, and it can be quite useful for exercises. You can find more about callout [here](https://quarto.org/docs/authoring/callouts.html).*

Example: Extend the above analysis to cover all genes.
:::
